Using spatially biased variables in ecosystem condition accounting with a GIS based workflow

### 2024-01-18 manuscript/HIAs.qmd
try to crop and mask EDM

---
title: Using spatially biased variables in ecosystem condition accounting with a GIS based workflow
subtitle: A Short Subtitle
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    affiliations: 
        - id: some-tech
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    attributes:
        corresponding: true
    note: This is the first author footnote.
  - name: Bob Security
    email: bob@example.com
    affiliations:
        - id: another-u
          name: Another University
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    note: |
      Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
  - name: Cat Memes
    email: cat@example.com
    affiliations:
        - ref: another-u
    note: Yet another author footnote.
  - name: Derek Zoolander
    email: derek@example.com
    affilations:
        - ref: some-tech
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - SEEA EA
  - alien species
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
execute:
  echo: false
  include: false
  eval: false
---

```{r setup, warning=F, message=F, include=F}
library(tidyverse)
library(knitr)
library(sf)

@@

```{r filepath}
# Set up conditional file paths
# The data is too big to be stored on GitHub
dir <- substr(getwd(), 1,2)

#pData <- ifelse(dir == "C:", 
#               "P:/41201785_okologisk_tilstand_2022_2023/data/",
#               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")
```

```{r}
# I already did some work to identify the relevant nature types

@@

# Getting the nature types that have 7FA
myVars <- "7FA"
naturetypes_summary2 <- naturetypes_summary %>%
  rowwise() %>%
  mutate(keepers = sum(c_across(
    all_of(myVars))>0, na.rm=T)) %>%
  filter(keepers >0,
         Ecosystem == "våtmark") %>%
  select(Nature_type, NiN_mainType, Year)

naturetypes_summary2$Nature_type

# Getting the nature types that have the ADSV variables
myVars <- c('7TK', '7SE', 'PRTK', 'PRSL')
naturetypes_summary3 <- naturetypes_summary %>%
  rowwise() %>%
  mutate(keepers = sum(c_across(
    all_of(myVars))>0, na.rm=T)) %>%
  filter(keepers >0,
         Ecosystem == "våtmark") %>%
  select(Nature_type, NiN_mainType, Year)

naturetypes_summary3$Nature_type
```

@@

```{r muniMaps, include=F, eval=F}
naturetypes <- sf::st_read(dsn = path) %>%
  # keep only wetlands
  filter(hoved.kosystem == "våtmark") %>%

  filter(naturtype %in% naturetypes_summary$Nature_type) %>%
  mutate(hovedøkosystem = case_match(hovedøkosystem,
                                  "naturligÅpneOmråderILavlandet" ~ "naturligÅpneOmråderUnderSkoggrensa",
                                  .default = hovedøkosystem),
         area = st_area(.)) %>%
  separate_rows(ninBeskrivelsesvariable, sep=",") %>%
  separate(col = ninBeskrivelsesvariable,
           into = c("NiN_variable_code", "NiN_variable_value"),
           sep = "_",
           remove=F) %>%
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) %>%
  filter(NiN_variable_code %in% myVars)

temp <- naturetypes %>%
    filter(hoved.kosystem == "våtmark")

>>>>

---
title: Using spatially biased variables in ecosystem condition accounting with a GIS based workflow
subtitle: A Short Subtitle
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    affiliations: 
        - id: some-tech
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    attributes:
        corresponding: true
    note: This is the first author footnote.
  - name: Bob Security
    email: bob@example.com
    affiliations:
        - id: another-u
          name: Another University
          department: Department Name
          address: Street Address
          city: City
          state: State
          postal-code: Postal Code
    note: |
      Another author footnote, this is a very long footnote and it should be a really long footnote. But this footnote is not yet sufficiently long enough to make two lines of footnote text.
  - name: Cat Memes
    email: cat@example.com
    affiliations:
        - ref: another-u
    note: Yet another author footnote.
  - name: Derek Zoolander
    email: derek@example.com
    affilations:
        - ref: some-tech
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - SEEA EA
  - alien species
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
editor: 
  markdown: 
    wrap: sentence
# remove this execute part to actually run the analysis
execute:
  echo: false
  include: false
  eval: false
---

```{r setup, warning=F, message=F, include=F}
library(tidyverse)
library(knitr)
library(sf)
library(tmap)
library(stars)
library(terra)
```

@@

```{r getRelevantNTs}
# I already did some work to identify the relevant nature types

@@

# Getting the nature types that have 7FA
myVars <- "7FA"

#naturetypes_summary2 <- naturetypes_summary %>%
#  rowwise() %>%
#  mutate(keepers = sum(c_across(
#    all_of(myVars))>0, na.rm=T)) %>%
#  filter(keepers >0,
#         Ecosystem == "våtmark") %>%
#  select(Nature_type, NiN_mainType, Year)
#
#naturetypes_summary2$Nature_type

# Getting the nature types that have the ADSV variables
myVars2 <- c('7TK', '7SE', 'PRTK', 'PRSL')
#naturetypes_summary3 <- naturetypes_summary %>%
#  rowwise() %>%
#  mutate(keepers = sum(c_across(
#    all_of(myVars2))>0, na.rm=T)) %>%
#  filter(keepers >0,
#         Ecosystem == "våtmark") %>%
#  select(Nature_type, NiN_mainType, Year)
#
#naturetypes_summary3$Nature_type
```

@@

```{r prepData}
# Set up conditional file paths
# The data is too big to be stored on GitHub
dir <- substr(getwd(), 1,2)

#pData <- ifelse(dir == "C:", 
#               "P:/41201785_okologisk_tilstand_2022_2023/data/",
#               "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/")

# Impart polygon data set
naturetypes <- sf::st_read(dsn = path)
# 117k polygons (2023)
# sometimes this functin struggles with Norwegian letters. Restart R and try again.

naturetypes <- naturetypes |>
  # keep only wetlands and our selected nature types
  filter(hovedøkosystem == "våtmark",
         naturtype %in% c(advsNTs, alienNTS)) |>
  # calculate the areas (m2) of the polygons
  mutate(area = SHAPE |> st_area()) |>
  # the variable codes and values are all in the same column
  separate_rows(ninBeskrivelsesvariable, sep=",") |>
  separate(col = ninBeskrivelsesvariable,
           into = c("NiN_variable_code", "NiN_variable_value"),
           sep = "_",
           remove=F) |>
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) |>
  filter(NiN_variable_code %in% c(myVars, myVars2)) |>
  select(id = identifikasjon_lokalId,
         municipality = kommunenummer,
         year = kartleggingsår,
         mosaic = mosaikk,
         quality = lokalitetskvalitet,
         biodiversity = naturmangfold,
         condition = tilstand,
         natureType = naturtype,
         variable = NiN_variable_code,
         value = NiN_variable_value,
         area)

# 13k obs.
#naturetypes |>
#  saveRDS("../data/surveyData.RDS")
#head(naturetypes)
```

```{r getNorwayOutline}
outline <- sf::read_sf("../data/outlineOfNorway_EPSG25833.shp")
```

```{r getMunicipalityOutine}
path <- paste0("../data/Basisdata_0000_Norge_25833_Kommuner_FGDB.gdb")

# find the correct layer
#st_layers(path)
muni <- sf::read_sf(path, options = "ENCODING=UTF8", layer = "kommune")
#st_crs(muni) #25833

# This returns some multisurfaces as well
table(st_geometry_type(muni))

# function to make sure that mutipolygons are returned 
ensure_multipolygons <- function(X) {
    tmp1 <- tempfile(fileext = ".gpkg")
    tmp2 <- tempfile(fileext = ".gpkg")
    st_write(X, tmp1)
    gdalUtilities::ogr2ogr(tmp1, tmp2, f = "GPKG", nlt = "MULTIPOLYGON")
    Y <- st_read(tmp2)
    st_sf(st_drop_geometry(X), geom = st_geometry(Y))
}

muni <- ensure_multipolygons(muni)
#> Writing layer `file34a3261c663332' to data source 
#>   `/tmp/RtmpkT1dSD/file34a3261c663332.gpkg' using driver `GPKG'
#> Writing 363 features with 13 fields and geometry type Unknown (any).
#> Reading layer `file34a3261c663332' from data source 
#>   `/tmp/RtmpkT1dSD/file34a32673f6fef.gpkg' 
#>   using driver `GPKG'
#> Simple feature collection with 363 features and 13 fields
#> Geometry type: MULTIPOLYGON
#> Dimension:     XY
#> Bounding box:  xmin: -99551.21 ymin: 6426048 xmax: 1121941 ymax: 7962744
#> Projected CRS: ETRS89 / UTM zone 33N
table(st_geometry_type(muni)) #OK

# code to show all municipalities
#tm_shape(muni)+
#  tm_polygons(col="kommunenummer")+
#  tm_layout(legend.outside = T)

# Transform to match the EDM (see below)
muni <- muni |>
  st_transform(25832)

# subset of the three taregt municipalities
muni3 <- muni |>
  filter(kommunenummer %in% c("3020", #Nordre Follo
                              "5001", # Trondheim
                              "5418"  # Målselv
                              ))
# To crop EDM, I need the three municipalities seprately. 
nf <- muni3 |>
  filter(kommunenummer == "3020")
tr <- muni3 |>
  filter(kommunenummer == "5001")
me <- muni3 |>
  filter(kommunenummer == "5418")

# Make map to show where the three municipalities are
#
#

# I need to intersect nature types with municipality
# and also to get the dat coverage polygon
tm_shape(muni3)+
  tm_polygons()+
  tm_facets(by = "kommunenummer")+
  tm_shape(naturetypes)+
  tm_polygons()
```

```{r getMireData}
file <- "P:/41201785_okologisk_tilstand_2022_2023/data/Myrmodell/myrmodell90pros.tif"
# This is a big TIF, so I some lazy evalyation on the starts proxy 
nf_mire  <- stars::read_stars(file, proxy=T) |>
  st_crop(nf) |>
  st_as_stars()

# Same result, different syntax:
#nf_mire  <- stars::read_stars(file, proxy=T)
#nf_mire <- nf_mire[nf]
#nf_mire <- st_as_stars(nf_mire)

# This is not working as it puts the cropped raster into the sea
#tmap_mode("view")
#tm_shape(st_downsample(nf_mire , 10))+
#  tm_raster(alpha=.8
#            )+
#  tm_shape(outline)+
#  tm_borders()

# trying terra
# convert municipal outline to vect via st
nf2 <- as(nf, "Spatial") |>
  terra::vect()

# crop and mask (very fast!)
nf_mire2 <- terra::rast(file) |>
  crop(nf2) |>
  terra::mask(nf2)

# Plot to check overlap
ggplot()+
  geom_spatraster(data = nf_mire2)+
  geom_spatvector(data = nf2,
                  fill = NA)
# The cropping and masking worked.

# I ilke the stars, sf and tmap combo better, so I return to stars
nf_mire3 <- nf_mire2 |>
  st_as_stars()

tmap_mode("plot")
tm_shape(st_downsample(nf_mire3 , 10))+
  tm_raster(alpha=.8
            )+
  tm_shape(outline)+
  tm_borders()

# Then I cut out the relevant part of the ecosystem map
nft <- nf |>
  terra::vect()

ras_crop <- terra::crop(ras_raw, shp)
ras_mask <- terra::mask(ras_crop, shp, touches = T)

nf_mire <- st_crop(EDM, st_bbox(nf))
tr_mire <- st_crop(EDM, tr)
me_mire <- st_crop(EDM, me)

  tm_shape(nf)+
  tm_borders(col="red", 
             lwd=3)+
tm_shape(st_downsample(nf_mire, 10))+
  tm_raster(col="myrmodell90pros.tif")

#tm_shape(st_downsample(EDM, 100))+
  #tm_raster(alpha=.3)+
